\clearemptydoublepage

\phantomsection
\addcontentsline{toc}{chapter}{Outline of the Thesis}

\begin{center}
	\huge{Outline of the Thesis}
\end{center}




%--------------------------------------------------------------------
\section*{Part I: Introduction and Theory}

\noindent {\scshape Chapter 1: Introduction}  \vspace{1mm}

\noindent  This chapter presents an overview of the thesis and its purpose. \\

\noindent {\scshape Chapter 2: Basics about USB}  \vspace{1mm}

\noindent  This chapter describes the basics about USB, how it is structured and how does it generally work.   \\

\noindent {\scshape Chapter 3: USB on Android}  \vspace{1mm}

\noindent  Chapter 3 introduces the reader to the USB API on Android. How can USB devices be enumerated and accessed, and how can the communication be initiated? \\

\noindent {\scshape Chapter 4: USB Mass storage class}  \vspace{1mm}

\noindent  This chapter gives an overview of the USB Mass storage class. The main goal is to understand SCSI commands to communicate with the device and read and write from and to its storage. \\

\noindent {\scshape Chapter 5: File systems}  \vspace{1mm}

\noindent  Chapter 5 discusses the approach of file system in general, and then gives a detailed description of how FAT32 from Microsoft works. \\

%--------------------------------------------------------------------
\section*{Part II: Implementation}

\noindent {\scshape Chapter 6: Purpose and Overview}  \vspace{1mm}

\noindent  This chapter presents the purpose and requirements for the implementation. It also provides a short overview of the implementation. \\

\noindent {\scshape Chapter 7: Inside the packages}  \vspace{1mm}

\noindent  Chapter 7 gives a deeper insight into the developed framework and its packages, classes and interaction between the important parts. \\

\section*{Part III: Quality Management}

\noindent {\scshape Chapter 8: Testing}  \vspace{1mm}

\noindent  This chapter presents the results of different tests regarding the developed framework on various devices. \\

\section*{Part IV: Results}

\noindent {\scshape Chapter 9: Summary}  \vspace{1mm}

\noindent  Short summary of the current status of the framework, and a conclusion of the thesis. \\

\noindent {\scshape Chapter 10: Outlook}  \vspace{1mm}

\noindent  This chapter gives some hints about further development and advices what future may hold. \\
